The Effectiveness of Testing, Vaccinations and Contact Restrictions for Containing the CoViD-19 Pandemic

### 2022-03-21 src/scientific_reports_submission/article.tex
Remove paper toc from supplmat toc, fix references in article.

\author[1, 2, +]{Janoś Gabler}
\author[3, +]{Tobias Raabe}
\author[1, +]{Klara Röhrl}
\author[2, 4, *, +]{Hans-Martin von Gaudecker}

@@

\affil[+]{These authors contributed equally to this work}

@@

Since early 2020, the CoViD-19 pandemic has presented an enormous challenge to humanity
on many dimensions. The development of highly effective vaccines holds the promise of
containment in the medium term. However, most countries find themselves many
months---and often years---away from reaching vaccination levels that would end the
pandemic or even protect the most vulnerable \cite{Mathieu2021}. In the meantime, it is
of utmost importance to employ an effective mix of strategies for containing the virus.
The most frequent initial response was a set of non-pharmaceutical interventions (NPIs)
to reduce contacts between individuals. While this has allowed some countries to sustain
equilibria with very low infection numbers (see \cite{Contreras2021} for a theoretical
equilibrium at low case numbers which is sustained with test-trace-and-isolate
policies), most have seen large fluctuations of infection rates over time. Containment
measures have become increasingly diverse and now include rapid testing, more nuanced
NPIs, and contact tracing. Neither these policies' effects nor the influence of seasonal
patterns or of more infectious virus strains are well understood in quantitative terms.

@@

At the core of our agent-based model---we review more literature in Supplementary
Material~B.1 \cite{Aleta2020,Hinch2021a}---are physical contacts between heterogeneous
agents (Figure~1a). Each contact between an infectious individual and somebody
susceptible to the disease bears the risk of transmitting the virus. Contacts occur in
up to four networks: Within the household, at work, at school, or in other settings
(leisure activities, grocery shopping, medical appointments, etc.). Some contacts recur
regularly, others occur at random. Empirical applications can take the population and
household structure from census data and the network-specific frequencies of contacts
from diary data measuring contacts before the pandemic \cite{Mossong2008,Hoang2019}.
Within each network, meeting frequencies depend on age and geographical location (see
Supplementary Material~A.4).

@@

In the model, susceptibility to contracting the SARS-CoV-2 virus is dependent on age
\cite{Davies2020,Goldstein2021}. A possible infection progresses as shown in Figure~1b.
We differentiate between an initial period of infection without being infectious or
showing symptoms, being infectious (presymptomatic or asymptomatic), showing symptoms,
requiring intensive care, and recovery or death similar to \cite{Aleta2020}. The
probabilities of transitioning between these states depend on age; their duration is
random and calibrated to medical literature (for a detailed description see
Supplementary Material~A.1). Conditional on the type of contact, infectiousness is
independent of age \cite{Jones2021}.

@@

We include two types of tests. Polymerase chain reaction (PCR) tests reveal whether an
individual is infected or not; there is no uncertainty to the result. PCR tests require
at least one day to be processed and there are aggregate capacity constraints. In
contrast, rapid antigen tests yield immediate results. Specificity and sensitivity of
these tests is set according to data analyzed in
\cite{Bruemmer2021,Scheiblauer2021,Ozcurumez2021}; sensitivity depends on the timing of
the test relative to the onset of infectiousness as in \cite{Smith2021}. We analyse
robustness to different assumptions in Supplementary Material~A.12. After a phase-in
period, all tests that are demanded will be performed. Figure~1c shows our model for
rapid test demand. Schools may require staff and students to be tested regularly. Rapid
tests may be offered by employers to on-site workers. Individuals may demand tests for
private reasons, which include having plans to meet other people, showing symptoms of
CoViD-19, and a household member having tested positively for the virus. We endow each
agent with an individual compliance parameter. This parameter determines whether she
takes up rapid tests. (Positive test results or symptoms lead most individuals to reduce
their contacts; this is why tests impact the actual contacts in Figure~1.)

Modelling a population of agents according to actual demographic characteristics means
that we can use a wide array of data to identify and calibrate the model's many
parameters (see Supplementary Material~A for a complete description). Contact diaries
yield pre-pandemic distributions of contacts for different contact types and their
assortativity by age group. Mobility data is used to model the evolution of work
contacts. School and daycare policies can be incorporated directly from official
directives. Administrative records on the number of tests, vaccinations by age and
region, and the prevalence of virus strains are generally available. Surveys may ask
about test offers, propensities to take them up, and past tests. Other studies'
estimates of the seasonality of infections can be incorporated directly. The remaining
parameters---most notably, these include infection probabilities by contact network and
the effects of some NPIs, see Supplementary Material~A.9---will be chosen numerically so
that the model matches features of the data (see \cite{McFadden1989} for the general
method). In our application, we keep the number of free parameters low in order to avoid
overfitting. The data features to be matched include official case numbers for each age
group and region, deaths, and the share of the B.1.1.7 strain.

@@

The model is applied to the second and third waves of the CoViD-19 pandemic in Germany,
covering the period mid-September 2020 to the end of May 2021. Figure~2a describes the
evolution of the pandemic and of its drivers. The black line in Figure~2a shows
officially recorded cases; the black line in Figure~2b the Oxford Response Stringency
Index \cite{Hale2020}, which tracks the tightness of non-pharmaceutical interventions.
The index is shown for illustration of the NPIs, we never use it directly. For
legibility reasons, we transform the index so that lower values represent higher levels
of restrictions. A value of zero means all measures incorporated in the index are turned
on. The value one represents the situation in mid-September, with restrictions on
gatherings and public events, masking requirements, but open schools and workplaces. In
the seven weeks between mid September and early November, cases increased by a factor of
ten. Restrictions were somewhat tightened in mid-October and again in early November.
New infections remained constant throughout November before rising again in December,
prompting the most stringent lockdown to this date. Schools and daycare centers were
closed, so were customer-facing businesses except for grocery and drug stores. From the
peak of the second wave just before Christmas until the trough in mid-February, newly
detected cases decreased by almost three quarters. The third wave in the spring of 2021
is associated with the B.1.1.7 (Alpha) strain, which became dominant in March
(Figure~2c)---B.1.617.2 (Delta) was first detected in Germany in April; at the end of
our simulation period it accounted for less than 5\% of cases. In early March, some NPIs
were relaxed; e.g., hairdressers and home improvement stores were allowed to open again
to the public. There were many changes in details of regulations afterwards, but they
did not change the overall stringency index.

@@

So why is rapid testing so effective? In order to shed more light on this question,
Figure~3d decomposes the difference in the scenario without rapid tests and the main
specification into the three channels for rapid tests. Tests at schools have the
smallest effect, which is largely explained by schools not operating at full capacity
during our period of study and the relatively small number of students. (18\% of our
population are in the education sector, e.g., pupils, teachers; 46\% are workers outside
the education sector.) Almost 40\% come from tests at the workplace. Despite the fact
that rapid tests for private reasons are phased in only in mid-March, they make up for
more than half of the total effect. The reason lies in the fact that a substantial share
of these tests is driven by an elevated probability to carry the virus, i.e., showing
symptoms of CoViD-19 or following up on a positive test of a household member. The
latter is essentially a form of contact tracing, which has been shown to be very
effective \cite{Contreras2021, Fetzer2021,Kretzschmar2020}. Indeed, a deeper analysis in
Supplementary Material~B.15 shows that the same amount of rapid tests administered
randomly in the population would not have been nearly as effective.

@@

\section*{Acknowledgements (not compulsory)}

@@

To include, in this order: \textbf{Accession codes} (where applicable);

>>>>

\author[1, 2]{Janoś Gabler}
\author[3]{Tobias Raabe}
\author[1]{Klara Röhrl}
\author[2, 4, *]{Hans-Martin von Gaudecker}

@@

@@

Since early 2020, the CoViD-19 pandemic has presented an enormous challenge to humanity
on many dimensions. The development of highly effective vaccines holds the promise of
containment in the medium term. However, most countries find themselves many
months---and often years---away from reaching vaccination levels that would end the
pandemic or even protect the most vulnerable \cite{Mathieu2021}. In the meantime, it is
of utmost importance to employ an effective mix of strategies for containing the virus.
The most frequent initial response was a set of non-pharmaceutical interventions (NPIs)
to reduce contacts between individuals. While this has allowed some countries to sustain
equilibria with very low infection numbers\textcolor{red}{---see \cite{Contreras2021}
for a theoretical equilibrium at low case numbers which is sustained with
test-trace-and-isolate policies---}, most have seen large fluctuations of infection
rates over time. Containment measures have become increasingly diverse and now include
rapid testing, more nuanced NPIs, and contact tracing. Neither these policies' effects
nor the influence of seasonal patterns or of more infectious virus strains are well
understood in quantitative terms.

@@

At the core of our agent-based model\cite{Aleta2020,Hinch2021a}---we review more
literature in Supplementary Material~B.1---are physical contacts between heterogeneous
agents (Figure~1a). Each contact between an infectious individual and somebody
susceptible to the disease bears the risk of transmitting the virus. Contacts occur in
up to four networks: Within the household, at work, at school, or in other settings
(leisure activities, grocery shopping, medical appointments, etc.). Some contacts recur
regularly, others occur at random. Empirical applications can take the population and
household structure from census data and the network-specific frequencies of contacts
from diary data measuring contacts before the pandemic \cite{Mossong2008,Hoang2019}.
Within each network, meeting frequencies depend on age and geographical location (see
Supplementary Material~A.4).

@@

In the model, susceptibility to contracting the SARS-CoV-2 virus is dependent on age
\cite{Davies2020,Goldstein2021}. A possible infection progresses as shown in Figure~1b.
We differentiate between an initial period of infection without being infectious or
showing symptoms, being infectious (presymptomatic or asymptomatic), showing symptoms,
requiring intensive care, and recovery or death (similar to \cite{Aleta2020}). The
probabilities of transitioning between these states depend on age; their duration is
random and calibrated to medical literature (for a detailed description see
Supplementary Material~A.1). Conditional on the type of contact, infectiousness is
independent of age \cite{Jones2021}.

@@

We include two types of tests. Polymerase chain reaction (PCR) tests reveal whether an
individual is infected or not; there is no uncertainty to the result. PCR tests require
at least one day to be processed and there are aggregate capacity constraints. In
contrast, rapid antigen tests yield immediate results. Specificity and sensitivity of
these tests is set according to data analyzed in
\cite{Bruemmer2021,Scheiblauer2021,Ozcurumez2021}; sensitivity depends on the timing of
the test relative to the onset of infectiousness as in \cite{Smith2021}. We analyse
robustness to different assumptions in Supplementary Material~B.12. After a phase-in
period, all tests that are demanded will be performed. Figure~1c shows our model for
rapid test demand. Schools may require staff and students to be tested regularly. Rapid
tests may be offered by employers to on-site workers. Individuals may demand tests for
private reasons, which include having plans to meet other people, showing symptoms of
CoViD-19, and a household member having tested positively for the virus. We endow each
agent with an individual compliance parameter. This parameter determines whether she
takes up rapid tests. \textcolor{red}{(Positive test results or symptoms lead most
individuals to reduce their contacts; this is why tests impact the actual contacts in
Figure~1.)}

Modelling a population of agents according to actual demographic characteristics means
that we can use a wide array of data to identify and calibrate the model's many
parameters \textcolor{red}{(see Supplementary Material~A for a complete description)}.
Contact diaries yield pre-pandemic distributions of contacts for different contact types
and their assortativity by age group. Mobility data is used to model the evolution of
work contacts. School and daycare policies can be incorporated directly from official
directives. Administrative records on the number of tests, vaccinations by age and
region, and the prevalence of virus strains are generally available. Surveys may ask
about test offers, propensities to take them up, and past tests. Other studies'
estimates of the seasonality of infections can be incorporated directly. The remaining
parameters---most notably, these include infection probabilities by contact network and
the effects of some NPIs, see Supplementary Material~A.9---will be chosen numerically so
that the model matches features of the data (see \cite{McFadden1989} for the general
method). In our application, we keep the number of free parameters low in order to avoid
overfitting. The data features to be matched include official case numbers for each age
group and region, deaths, and the share of the B.1.1.7 strain.

@@

The model is applied to the second and third waves of the CoViD-19 pandemic in Germany,
covering the period mid-September 2020 to the end of May 2021. Figure~2 describes the
evolution of the pandemic and of its drivers. The black line in Figure~2a shows
officially recorded cases; the black line in Figure~2b the Oxford Response Stringency
Index \cite{Hale2020}, which tracks the tightness of non-pharmaceutical interventions.
The index is shown for illustration of the NPIs, we never use it directly. For
legibility reasons, we transform the index so that lower values represent higher levels
of restrictions. A value of zero means all measures incorporated in the index are turned
on. The value one represents the situation in mid-September, with restrictions on
gatherings and public events, masking requirements, but open schools and workplaces. In
the seven weeks between mid September and early November, cases increased by a factor of
ten. Restrictions were somewhat tightened in mid-October and again in early November.
New infections remained constant throughout November before rising again in December,
prompting the most stringent lockdown to this date. Schools and daycare centers were
closed, so were customer-facing businesses except for grocery and drug stores. From the
peak of the second wave just before Christmas until the trough in mid-February, newly
detected cases decreased by almost three quarters. The third wave in the spring of 2021
is associated with the B.1.1.7 (Alpha) strain, which became dominant in March
(Figure~2c). \textcolor{red}{B.1.617.2 (Delta) was first detected in Germany in April;
at the end of our simulation period it accounted for less than 5\% of cases.} In early
March, some NPIs were relaxed; e.g., hairdressers and home improvement stores were
allowed to open again to the public. There were many changes in details of regulations
afterwards, but they did not change the overall stringency index.

@@

So why is rapid testing so effective? In order to shed more light on this question,
Figure~3d decomposes the difference in the scenario without rapid tests and the main
specification into the three channels for rapid tests. Tests at schools have the
smallest effect, which is largely explained by schools not operating at full capacity
during our period of study and the relatively small number of students.
\textcolor{red}{18\% of our population are in the education sector, e.g., pupils,
teachers; 46\% are workers outside the education sector.} Almost 40\% come from tests at
the workplace. Despite the fact that rapid tests for private reasons are phased in only
in mid-March, they make up for more than half of the total effect. The reason lies in
the fact that a substantial share of these tests is driven by an elevated probability to
carry the virus, i.e., showing symptoms of CoViD-19 or following up on a positive test
of a household member. The latter is essentially a form of contact tracing, which has
been shown to be very effective \cite{Contreras2021, Fetzer2021,Kretzschmar2020}.
Indeed, a deeper analysis in Supplementary Material~B.15 shows that the same amount of
rapid tests administered randomly in the population would not have been nearly as
effective.

@@

\section*{Acknowledgements}

@@
